---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# memoria

  <!-- badges: start -->
[![CRAN_Release_Badge](http://www.r-pkg.org/badges/version-ago/memoria)](https://CRAN.R-project.org/package=memoria)
[![CRAN_Download_Badge](http://cranlogs.r-pkg.org/badges/grand-total/memoria)](https://CRAN.R-project.org/package=memoria)
[![R-CMD-check](https://github.com/blasbenito/memoria/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/blasbenito/memoria/actions/workflows/R-CMD-check.yaml)
 <!-- badges: end -->

The **memoria** package quantifies ecological memory in long time-series involving environmental drivers and biotic responses, including palaeoecological datasets. It uses Random Forest models to measure the relative importance of endogenous memory (effect of past response values), exogenous memory (effect of past driver values), and concurrent effects (synchronic driver influence).

## Installation

Install from CRAN:

```{r, eval = FALSE}
install.packages("memoria")
```

Or the development version from GitHub:

```{r, eval = FALSE}
# install.packages("devtools")
devtools::install_github("BlasBenito/memoria")
```

## Quick Example

This minimal example uses built-in data to demonstrate the core workflow:

```{r example, message = FALSE, warning = FALSE}
library(memoria)

# Load pre-prepared lagged data
data(palaeodataLagged)

# Compute ecological memory pattern
memory_output <- computeMemory(
  lagged.data = palaeodataLagged,
  drivers = c(
    "climate.temperatureAverage", 
    "climate.rainfallAverage"
    ),
  response = "Response",
  add.random = TRUE,
  random.mode = "autocorrelated",
  repetitions = 100
)

# Plot the memory pattern
x <- plotMemory(memory_output)
```

The plot shows:

- **Violet curve**: Endogenous memory (effect of past pollen abundance)
- **Blue/green curves**: Exogenous memory (effect of past climate variables)
- **Yellow band**: Random benchmark for significance testing
- **Lag 0**: Concurrent effect of drivers

## Learn More

- **[Quick Start Guide](https://blasbenito.github.io/memoria/articles/quick_start.html)**: Full workflow from raw data to memory features
- **[Using memoria](https://blasbenito.github.io/memoria/articles/using_memoria.html)**: Detailed explanation of methods and theory
- **[Function Reference](https://blasbenito.github.io/memoria/reference/index.html)**: Documentation for all exported functions

## How to cite memoria

If you use **memoria** in your research, please cite both the package and the associated paper:

**Paper:**

> Benito, B.M., Gil-Romera, G., & Birks, H.J.B. (2020). Ecological memory at millennial time-scales: the importance of data constraints, species longevity, and niche features. *Ecography*, 43(1), 1-10. doi: [10.1111/ecog.04772](https://doi.org/10.1111/ecog.04772)

**R package:**

> Benito, B.M. (2025). memoria: Quantifying Ecological Memory in Palaeoecological Datasets and Other Long Time-Series. R package version 1.1.0. doi: [10.32614/CRAN.package.memoria](https://doi.org/10.32614/CRAN.package.memoria)
